Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-13 paper/ms.tex
Update title, abstract, and keywords

\title{Rising from the Ashes: How the Milky Way Got Its Scars}

@@

\begin{abstract}
    The elemental abundance distribution of stars encodes the history of the gas-phase abundance in the Milky Way. Without a large, unbiased sample of highly precise stellar ages, the exact timing and nature of this history must be \textit{inferred} from the abundances. In the two-dimensional plane of \alphaFe-\FeH, it is now clear that two separate populations exist -- the low-$\alpha$ and high-$\alpha$ sequences. Structure in the elemental abundance distribution can arise from many processes -- proposals include specific gas infall scenarios, radial migration, high-redshift clump formation, and various effects associated with galaxy mergers, among others. In this work, we demonstrate another possible avenue for structure formation with clear observational predictions. In this scenario, the Galaxy underwent a starburst followed by a brief (hundreds of Myr) quiescent phase -- i.e., the Galaxy underwent a post-starburst rejuvenation sequence at $z\sim2$. A natural consequence of the quiescent phase is that stars in the valley of the bimodality do not form because: (1) the absence of enrichment from high-mass stars leads to a rapid reduction in \alphaFe{}, and (2) any time the gas spends in the abundance valley is deemphasized in the present day distribution because the star formation rate is lower. With a set of idealized merger simulations, we demonstrate the feasibility of this proposal. This ``phoenix hypothesis'' predicts a $\sim300\,\Myr$ gap in stellar ages at a fixed \FeH{} and that stars which form directly after this gap would have lower \alphaFe{} than stars which form slightly ($\sim1\,\Gyr$) later.
  \end{abstract}

\keywords{Classical Novae (251) --- Ultraviolet astronomy(1736) --- History of astronomy(1868) --- Interdisciplinary astronomy(804)}

>>>>

\title{Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences}

@@

\begin{abstract}
    The elemental abundance distribution of stars encodes the history of the gas-phase abundance in the Milky Way. Without a large, unbiased sample of highly precise stellar ages, the exact timing and nature of this history must be \textit{inferred} from the abundances. In the two-dimensional plane of \alphaFe{}-\FeH{}, it is now clear that two separate populations exist -- the low-$\alpha$ and high-$\alpha$ sequences. \edit1{We propose that a brief ($\sim300\,\Myr$) halt in star formation within a narrow metallicity bin can lead to a bimodal \alphaFe{} distribution at that metallicity, assuming a rapidly declining gas phase \alphaFe{}. Using simulations of an idealized setup of a high-$z$ galaxy merger, we show that the merger with the Gaia-Sausage-Enceladus satellite at $z\sim2$ is one possible way to trigger such a gap in the Milky Way. Our proposal does not rely on a merger to induce this age gap. We predict a $\sim300\,\Myr$ gap in stellar ages at a fixed \FeH{} where the $\alpha$-bimodality is prominent ($\FeH\lesssim-0.2$).}
\end{abstract}

\keywords{Milky Way disk (1050) --- Milky Way Galaxy physics (1056) --- Milky Way formation (1053), Hydrodynamical simulations (767) --- Post-starburst galaxies (2176) --- Galaxies (573) --- Starburst galaxies (1570)}
